---
title: "Influenza Differential Expression Analysis"
author: "Alperen Korkmaz"
format: 
  html:
    theme: cosmo          
    embed-resources: true
---

# Generating a volcano plot from an expression data set obtained from the GEO database

Installing needed packages and libraries.

```{r}

library(tidyverse)
library(DESeq2)
library(Biobase)
library(GEOquery)
library(janitor)
library(EnhancedVolcano)
library(limma)
library(knitr)
library(rlang)
library(data.table)
```

We import the dataset and the supporting files.

```{r}

GSE68849 <- getGEO("GSE68849", GSEMatrix = TRUE, destdir = ".")
supp_files <- getGEOSuppFiles("GSE68849")
#
eset68849 <- GSE68849[[1]]
```

Tidying the data.

The GEO dataset contains various types of information; we have specific functions to extract each of them individually: exprs() .Retrieves the expression dataset from the eSet object. pData() .Retrieves clinical/metadata regarding the samples (phenotype data); information on experimental phenotypes recorded in eSet and ExpressionSet-derived classes. fData() .Retrieves information on features recorded in eSet-derived classes (feature data); extracts gene and annotation information related to the probes. experimentData() and annotation() .Contain general information—such as who conducted the experiment, the publication date, the PubMed ID, and the laboratory protocols used. These are important for reporting purposes rather than for data analysis.

Getting clinical/metadata.

```{r}
p_data <- pData(eset68849)
head(p_data)
```

Getting gene expression data. And we are organizing it so that we can process it and then we take the log2 for the computation.

```{r}
exprs_data <- fread("C:/Users/alper/Desktop/Quarto Influenza A/GSE68849/GSE68849_non-normalized.txt.gz", data.table = FALSE)

rownames(exprs_data) <- make.names(exprs_data[, 1], unique = TRUE)

exprs_data <- exprs_data[, -1]


signal_column <- grep("AVG_Signal", colnames(exprs_data))


exprs_clean <- exprs_data[, signal_column]

colnames(exprs_clean) <- gsub(".AVG_Signal", "", colnames(exprs_clean))

found_column <- apply(p_data, 2, function(x) any(grepl("5522887032", x, ignore.case = TRUE)))
names(found_column)[found_column]
colnames(p_data)

colnames(exprs_clean) <- rownames(p_data)

exprs_log2 <- log2(exprs_clean + 1)
  
exprs_log2 <- exprs_clean
  
```

Getting gene information data.

```{r}
gen_data <- fData(eset68849)
head(gen_data)
```

For this project I didn't need experimental data and annotation data.

The expression data lacks probe IDs; We need to add this information to it. The gene data contains this information, and since both data frames are of the same length, they correspond to one another. Therefore, we will extract the probe IDs and gene names from the gene data and transfer them to the expression data. Also we simplify it the gene data.

```{r}
exprs_log2 <- exprs_log2 |> 
  as.data.frame() |> 
  rownames_to_column(var = "Prob_ID") |> 
  as_tibble()

gen_data <- gen_data |> 
  select(c("ID","ILMN_Gene"))

exprs_log2 <- left_join(exprs_log2, gen_data, by = c("Prob_ID" = "ID")) |>  
  relocate(`ILMN_Gene`, .after = Prob_ID)

```

We will clean the NAs in the expression data.

```{r}
exprs_log2 <- exprs_log2 |> 
  drop_na()
```

We want to create a Volcano Plot. Initially, we apply the log2 of the data. Now we need to calculate fold change and p value. For these we will use the limma library. But before the using the library we must turn our data frame to matrix. And we must eliminate duplicate copies of the same genes.

```{r}
expression_data_single_genes <- exprs_log2 |> 
  group_by(`ILMN_Gene`) |> 
  summarize(across(where(is.numeric), mean, na.rm = TRUE)) |> #We calculate the average of the data from the same genes.
  ungroup()

expression_data_single_genes <- as.data.frame(expression_data_single_genes)

rownames(expression_data_single_genes) <- expression_data_single_genes$`ILMN_Gene`
expr_matrix <- expression_data_single_genes|> 
  select(-`ILMN_Gene`) |> 
  as.matrix()

```

Now we need to determine which data belong to the patient and which to the control group. If we look up the clinical data.

```{r}
head(p_data, 10) |> 
  knitr::kable(caption = "Örneklerin Klinik Dağılımı")
```

The control comes first then comes patient and again control and then again patient. Like ABABABABAB

```{r}
groups_ <- factor(rep(c("control","patient"), 5))
#Crating the model matrix.
groups_ <- factor(groups_, levels = c("control", "patient"))
design <- model.matrix(~0 + groups_)
colnames(design) <- levels(groups_)

```

And now building and comparing the model to Limma

```{r}
fit <- lmFit(expr_matrix, design)
```

Showing the which groups to the difference will be calculated.

```{r}
contrast_matrix <- makeContrasts(patient_vs_control = patient - control, levels = design)
fit2 <- contrasts.fit(fit, contrast_matrix)
```

And adding the eBayes function for increase the statistical reliability.

```{r}
fit2 <- eBayes(fit2)
```

For the last one we getting result table.

```{r}
result_table <- topTable(fit2, adjust = "fdr", number = Inf)
```

We have all the data we need for Volcano plot. We use EnhancedVolcano library to create plot more easily.

Let's enter the necessary data to function.

```{r, fig.width=10, fig.height=8}
EnhancedVolcano(result_table,
                lab = rownames(result_table),
                x = 'logFC',
                y = 'P.Value',
                title = 'Influenza A Gene Expression Change',
                subtitle = 'Changes in gene expression in individuals infected with Influenza A \n virus compared to healthy individuals.',
                axisLabSize = 12,
                legendLabSize = 12,
                pCutoff = 0.05,
                FCcutoff = 1.0,
                pointSize = 3.0,
                labSize = 5.0,
                legendPosition = 'bottom'
)
```

And the result. Result shows the increase of the IFNA family. IFNA gene family is produces potent antiviral and immunomodulatory proteins that body cells secrete in response to viral infections, cancer, and immune system signals. Here the patients has Influenza A virus. This is explains the accuracy of our table.
